---
title: "Lecture 02 exercise solution"
author: "Becky Lin"
output: 
  html_document:
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Predict the outcome of the following:
```
c(1, FALSE)
c("a",1)
c(list(1),"a")
c(TRUE,1)
c(FALSE,2L)
typeof(c(1,2,3, NA))
is.infinite(c(1,Inf,-Inf, NaN))
is.nan(c(1,Inf,-Inf, NaN))
typeof(c("hello",9L, TRUE)) 
```
*Solution* You can see what they return by typing each line into the 
R console. The key point is that combining different data types coerces
the less flexible type to the more flexible type. 
The order of preference is: character $>$double $>$ integer $>$logical.

```{r}
c(1, FALSE)
c("a",1)
c(list(1),"a")
c(TRUE,1)
c(FALSE,2L)
typeof(c(1,2,3, NA))
is.infinite(c(1,Inf,-Inf, NaN))
is.nan(c(1,Inf,-Inf, NaN))
typeof(c("hello",9L, TRUE)) 
```

2. Explain the result

```{r}
1 == "1"
-1 < FALSE
"one" < 2 
```

**Solution:** Comparing different data types coerces the less flexible type to the more flexible type. The order of preference is: character $>$double $>$ integer $>$logical. So the first one is equivalent to compare "1"=="1" which leads to TRUE, for the second one, FALSE coerces to 0 then we compare $-1< 0$ which is TRUE.For the last one, essentially we compare "one" and "2", you will get FALSE as `utf8ToInt("o")` = 111 and `utf8toInt("2")` = 50.

3. If `vv <- list(a=1,b=2)`, why doesn't `as.vector(vv)`
work to coerce `vv` into an atomic vector?

**Solution:** Because `vv` is already a vector. 
```{r}
vv <- list(a=1,b=2)
is.vector(vv)
unlist(vv)
```

4. What do `dim()`, `nrow()` and `ncol()` return when applied to a 1-dimensional vector? What about `NROW()` or `NCOL()`?

**Solution:** Try `x <- c(1,2,3); dim(x); nrow(x); ncol(x)` to see that
each returns NULL, because a 1-d vector has no dim attribute. NROW and 
NCOL treat a vector as a 1-column matrix.

```{r}
x <- c(1,2,3) 
dim(x) 
nrow(x) 
ncol(x)
NROW(x)
NCOL(x)
```

5. What is `dim(cbind(A,A))` if `A = matrix(1:4,2,2)`?

**Solution:** we combine two 2 by 2 matrix by columns, so it results in a 2 by 4 matrix. The `dim()` will give: 2 4.
```{r}
A = matrix(1:4,2,2)
cbind(A,A)
dim(cbind(A,A))
```


6. What do the following return? Understand why. 

```
TRUE | FALSE
c(TRUE,TRUE) & c(FALSE,TRUE)
c(TRUE,TRUE) && c(FALSE,TRUE)
```

**Solution:** The key here is that & is vectorized while && only checks the first element of the two vectors.

```{r}
TRUE | FALSE
c(TRUE,TRUE) & c(FALSE,TRUE)
c(TRUE,TRUE) && c(FALSE,TRUE)
```

Remember the following truth tables

| condition1 (X) | condition2 (Y) | Not X (~X) | X AND Y (X && Y) | X OR Y (X\|\|Y) |
|-----------------| -----------------|------------| -----------------|---------------|
|FALSE |FALSE | TRUE| FALSE |FALSE|
|FALSE  |TRUE | TRUE| FALSE |TRUE|
|TRUE |FALSE | FALSE| FALSE |TRUE|
|TRUE |TRUE | FALSE| TRUE |TRUE|



 

7. What sort of object does `table()` return? What is its type? What attributes does it have? How does the dimensionality change as you tabulate more variables?

*Solution* An array with class attribute "table". The array is of type int.
The dimension increases with each variable. Tabulating 2 variables returns
a 2-d array (matrix), 3 variables returns a 3-d array, etc. Try the 
following in R

```
tt <- table(c(1,1,2,2,3,3),c(1,1,1,3,3,3),c(1,1,1,1,2,2)) 
tt
str(tt)
```


8. What happens to a factor when you modify its levels?
How do f2 and f3 differ from f1?

```
f1 <- factor(letters)
levels(f1) <- rev(levels(f1))
f2 <- rev(factor(letters))
f3 <- factor(letters, levels = rev(letters))
```

*Solution* The labels on the underlying integer values change, but
the integer values themselves don't. Try the following:

```{r}
f1 <- factor(letters)
f1
str(f1)
levels(f1) <- rev(levels(f1))
f1
str(f1)
f2 <- rev(factor(letters))
f2
str(f2)
f3 <- factor(letters, levels = rev(letters))
f3
str(f3)
data.frame(f1,as.numeric(f1),f2,as.numeric(f2),f3,as.numeric(f3))
```

You can see that f2 and f1 print the same but the integer-to-label
correspondence is backwards. Conversely, f3 and f3 have the same 
integer-to-label correspondence, but the labels are in reverse order.



9. Fix each of the following common data frame subsetting errors:
```
mtcars[mtcars$cyl = 4, ]
mtcars[-1:4, ]
mtcars[mtcars$cyl <= 5]
mtcars[mtcars$cyl == 4 | 6, ]
```

*Solution* 

```
mtcars[mtcars$cyl == 4,]
mtcars[-c(1:4),]
mtcars[mtcars$cyl <= 5,]
mtcars[mtcars$cyl ==4 | mtcars$cyl == 6, ]
```

10. Consider the following data frame and tibble. Suppose
you want to extract the number from the first row and first column and 
multiply this number by the vector `1:3`. How would
you do this for the data frame? How would you do this
for the tibble?
```
dd <- data.frame(x=1:3,y=c("a","b","c))
tt <- as.tibble(dd)
```


*Solution*:

```{r}
library(tidyverse)
dd <- data.frame(x=1:3,y=c("a","b","c"))
tt <- as.tibble(dd) 
dd[1,1]*(1:3)
as.numeric(tt[1,1])*(1:3)
```


11. Read the help file `help("%in%")`. Copy the following data frame into R.
```
students <- data.frame(
           major=c("Stat","Stat","Econ","Bus","Phys","ActSci","ActSci"),
           year=c(3,4,4,5,3,3,4))
```
Use `%in%` to create a logical vector of students who major in Stat
or ActSci and use this to subset `students` to Stat or ActSci majors.

*Solution*

```{r}
students <- data.frame(
           major=c("Stat","Stat","Econ","Bus","Phys","ActSci","ActSci"),
           year=c(3,4,4,5,3,3,4))
students[students$major %in% c("Stat","ActSci"),]
```
