---
title: "Lecture 05 exercises solution"
author: "Becky Lin"
output: 
  html_document:
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## I. Recursive partitioning by forward selection

* In this exercise you will write a version of 
recursive partitioning by forward selection
following Algorithm 1 of Friedman (1991), with 
a few notational changes that will make 
the transition to the MARS forward selection 
algorithm easier.

* The following code gives the framework for the function.
You are asked to write several of the support 
functions described below the code chunk.



```{r,attr.source='.numberLines'}
recpart_fwd <- function(y,x,Mmax){
  #---------------------------------------------------
  # Error checking for Mmax:if Mmax<2, throw a warning
  #                         and set it to 2
  #
  if(Mmax<2) {
    warning("Input Mmax must be >= 2; setting to 2")
    Mmax <- 2
  }
  #---------------------------------------------------
  # Initialize:
  N <- length(y) # sample size
  n <- ncol(x) # number of predictors = number of X
  # B: a data frame with optimal basis function as columns
  B <- init_B(N,Mmax) # initiate B dataframe
  # splits: a data frame records all the optimal (m, v, t):
  # m: parent basis func to split, v: splitting var, t: splitting point
  splits <- data.frame(m=rep(NA,Mmax),v=rep(NA,Mmax),t=rep(NA,Mmax))
  #
  #---------------------------------------------------
  # Looping for forward selection:
  for(M in 1:Mmax) { # contrast to indexing 2...Mmax in Friedman
    lof_best <- Inf
    for(m in 1:M) { # choose a basis function to split
      for(v in 1:n){ # select a variable to split on
        tt <- split_points(x[,v],B[,m]) # Exercise: write split_points() 
        for(t in tt) { 
          Bnew <- data.frame(B[,(1:M)[-m]], # drop m-th col: B[,-m]
                             # replace parent B[,m] with Btem1,Btem2
                             Btem1=B[,m]*(x[,v]>t), 
                             Btem2=B[,m]*(x[,v]<=t)) 
          gdat <- data.frame(y=y,Bnew)
          lof <- LOF(y~.,gdat) #  Use your LOF() from week 4
          if(lof < lof_best) { 
            lof_best <- lof
            splits[M,] <- c(m,v,t) 
          } # end if
        } # end loop over splits
      } # end loop over variables
    } # end loop over basis functions to split
    # save optimal (m, v, t) and update basis functions
    mstar <- splits[M,1]; vstar <- splits[M,2]; tstar <- splits[M,3]
    cat("[Info] best (m,v,t,lof): (",mstar,vstar,tstar,lof_best,")\n")
    B[,M+1] <- B[,mstar]*(x[,vstar]<=tstar) # 
    B[,mstar] <- B[,mstar]*(x[,vstar]>tstar)
  } # end loop over M
  return(list(B=B,splits=splits))
}

LOF <- function(form,data) {
  ff <- lm(form,data)
  return(sum(residuals(ff)^2))
}

#-------------------------------------
# Exercise Solutions are from here down

init_B <- function(N,Mmax) {
  # Input: N- # of rows; Mmax: # of basis funcs
  # output: a N by (Mmax+1) dataframe
  B <- data.frame( matrix(NA,nrow=N,ncol=(Mmax+1)) )
  B[,1] <- 1 # first column for intercept: B0
  names(B) <- c("B0",paste0("B",1:Mmax))
  return(B)
}

split_points <- function(xv,Bm) {
  # input: xv: a variable xv to split
  #        Bm: a parent basis func to split
  # output: feasible splitting points
  #
  out <- sort(unique(xv[Bm>0]))
  return(out[-length(out)])
}
```

Exercises:

1. Write a code snippet that checks 
whether the input `Mmax` is greater than or equal to 2. If not, 
throw a warning and set Mmax to 2.
2. Write the function `init_B()`, which takes the sample 
size `N` and max number of basis functions `Mmax` as input
and returns a data frame with 1's in the first column, NAs in
all others, and column names B0,B1,...,BMmax.
3. Write `split_points()`. This one is different from the 
one you wrote in the week 4 exercises. Here we take predictor
variable $x_v$, and basis function $B_m$ as input. The eligible split 
points are $x_v$ values such that basis function $B_m(x)>0$.
Return ordered unique values, and remember to exclude the 
largest possible value.
4. Test that `recpart_fwd(y,x,Mmax=2)`, with your functions from 
the exercises, runs on the following test data. 
These are the same as in the week 4 exercises.
See if you can find these first two splits on the drawing from week 4: 

[image](exrecpart.png)

```{r}
set.seed(123); n <- 10
x <- data.frame(x1=rnorm(n),x2=rnorm(n))
y <- rnorm(n)
rp2 <- recpart_fwd(y,x,Mmax=2)
rp2

# try Mmax=9
rp <- recpart_fwd(y,x,Mmax=9)
rp
```


* You will create further tests of your implementation in lab 3.

